\documentclass[10pt,twoside, svgnames]{book}
\usepackage{amsfonts, amsmath}
\usepackage{amssymb, url}
\usepackage{latexsym}
\usepackage{mathrsfs}
\newenvironment{thm}[1]{\begin{trivlist}
\item[{\bfseries #1}]}{\end{trivlist}}

\usepackage{csquotes}
\usepackage[%
    backend=biber,
    style=numeric-comp,
    citetracker=true,
    pagetracker=true,
    hyperref=true,
    backref=true,
    firstinits=true,
    bibencoding=utf8
    ]{biblatex}

%\newif\iffootnote
%\let\Footnote\footnote
%\renewcommand\footnote[1]{\begingroup\footnotetrue\Footnote{#1}\endgroup}

\makeatletter
\renewbibmacro*{cite:comp}{%
  \xdef\cbx@citekey{\thefield{entrykey}}%
  \addtocounter{cbx@tempcntb}{1}%
  \ifciteseen
    {}
    {\csnumgdef{cbx@instcount\cbx@citekey}{-100}}%  
  \ifsamepage{\value{instcount}}{\number\csuse{cbx@instcount\cbx@citekey}}%
    {\usebibmacro{author}, }
    {\renewcommand{\@makefntext}[1]{\noindent\normalfont##1}%
     \usebibmacro{author},
     \footnotetext{
       \printtext[labelnumberwidth]{%
         \printfield{labelprefix}%
         \printfield{labelnumber}}%
       \addspace
       \renewcommand{\newunitpunct}{\addcomma\addspace}
       \bibfootnotewrapper{%\fullcite{\thefield{entrykey}}%
         \printfield{labelname}\usebibmacro{author}\newunit\printfield{year}\newunit\printfield{title}\addperiod}}}%
   \csnumgdef{cbx@instcount\cbx@citekey}{\value{instcount}}%
   \iffieldundef{shorthand}
    {\ifbool{bbx:subentry}
       {\iffieldundef{entrysetcount}
          {\usebibmacro{cite:comp:comp}}
          {\usebibmacro{cite:comp:inset}}}
       {\usebibmacro{cite:comp:comp}}}
    {\usebibmacro{cite:comp:shand}}}
\makeatletter

\addbibresource{Local-TME-EMT.bib}

\usepackage{placeins}
\usepackage{epic}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pstricks}
\usepackage{pst-plot}

\usepackage[colorlinks=true, pdftex]{hyperref}
\hypersetup{citecolor=blue, linkcolor=blue}
\renewcommand{\sectionmark}[1]{\markright{\thesection\ #1}}%rappel du titre de la section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{fancyhdr}
\pagestyle{fancy} 
 % ceci permet d'avoir les noms de chapitre et section en minuscules
\renewcommand{\sectionmark}[1]{\markright{\thesection\ #1}}
  %rappel du titre de la section
\renewcommand{\subsectionmark}[1]{\markright{\thesubsection\ #1}}
  %rappel du titre de la section

\fancyhf{} \fancyhead[LE,RO]{\bfseries\thepage}
  %met en gras les entetes definies gauches paires et droite impaires
\fancyhead[LO]{\bfseries{\sc \rightmark}}%definit l'entete gauche impaire
\fancyhead[RE]{\bfseries{\small \sc \leftmark}}%definit l'entete droite paire
%%%% Les trois suivantes sont dans le fichier tex !!
\fancyfoot[LE,RO]{{\small \today}}%\copyright\ 2005}}
      %definit le pied de page gauche pair et droit impair 
%%%%%
\renewcommand{\headrulewidth}{0.5pt} %definit une ligne a la suite de l'entete
\renewcommand{\footrulewidth}{0.5pt} 
  %definit une ligne a la suite du pied de page
\addtolength{\headheight}{0.5pt} % espace pour le filet
\fancypagestyle{plain}{ %pages de tetes de chapitre :
  % redefinition du style plain pour avoir un debut de chapitre different
    \fancyhead{} %supprime l'entete
    \fancyfoot{} %supprime le pied de page
    \renewcommand{\headrulewidth}{0pt} %le filet haut
    \renewcommand{\footrulewidth}{0pt}} %le filet bas
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fancyfoot[LO]{{\sl The TME-EMT project, 2021}}
      %definit le pied de page gauche impair
\fancyfoot[RE]{\date}%definit le pied de page droit pair
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\nonstopmode

\title{{\textsc{\Huge The TME-EMT Project
         \\---\\[1ex]
      The PDF Notes\\[2ex]
       ---}}\\[1ex]
       %Version 2\\
       {\huge Th\'eorie Multiplicative Explicite des nombres
         \\ ---\\[1ex]
         Explicit Multiplicative number Theory
         }
     \\[1ex]
       ---}

\author{Contributors:\\ {\sl Olivier Bordell\`es, Pierre Dusart,
    Charles Greathouse}\\ {\sl Harald Helfgott, Pieter
  Moree, Akhilesh P.,  Olivier Ramar\'e} \\ {\sl \& Enrique Trevi\~no}}

\begin{document}
\frontmatter
\maketitle
%\relax{}
%\relax

%\hypersetup{pageanchor=true}

\mainmatter
\chapter*{Introduction}

Fully explicit results in multiplicative number theory are often
scattered through the litterature. The aim of this site is to present
an annoted bibliography in order to keep track of the current
knowledge. By the way, the acronym TME-EMT stands for 
\begin{center}
    Th\'eorie Multiplicative Explicite des nombres  

    ---

    Explicit
    Multiplicative number Theory 
\end{center}

Being up-to date is a difficulty, so please do not consider this site
as presenting the best available, but as a first line on which to
strengthen more specialized inquiries.

  {\small \tableofcontents\par}

\vfill
\section*{Notation}

Notation is standard, except maybe for the following one: we write
$f=O^*(g)$ to say that $|f|\le g$. This is simply a Landau-bigO symbol
with an implied constant equal to one. Furthermore, the letter p
always denotes a prime variable.

\part{Averages of arithmetical functions}

\input{Art01.tex}

\input{Art02.tex}
\input{Art10.tex}
\input{Art12.tex}

\part{Exact computations}

\input{Art03.tex}

\input{Art04.tex}

\part{General analytical tools}

\input{Art16.tex}

\input{Art17.tex}

\part{Exponential sums / points close to curves}

\input{Art05.tex}

\input{Art11.tex}


\part{Size of $L(1,\chi)$ and character sums }

\input{Art07.tex}
%\end{document}
\input{Art15.tex}

\part{Zeros and zero-free regions}

\input{Art06.tex}
\input{Art08.tex}

\part{Sieve and short interval results }

\input{Art09.tex}
\input{Art14.tex}

\part{Analytic Number Theory in Number Fields}

\input{Art18.tex}

\part{Applications}

\input{Art13.tex}
\input{Art19.tex}

\part{Development}

\chapter{README}

\section{How to write}

This part is technical and destined at the development team only.

\begin{enumerate}
\item Sections are coded via \texttt{<div
    class="section">1. Title</div>}. The numbering is done by hand.
\item Theorems, Lemmas, Propositions are coded via
  \begin{verbatim}
<span class="THM">Theorem (1986)</span>
<blockquote class="outer-thm">
<div class="thm">
    Every zero $\rho$ of $\zeta$ that have a
    real part between 0 and 1 and an imaginary part not more, 
    in absolute value, than $T_0=545439823$ are in fact on
    the critical line, i.e. satisfy $\Re \rho=1/2$.
</div></blockquote> 
\end{verbatim}
Note that the part \verb|</div></blockquote>|
should be on a single line.
  
\item Mathematics are entered latex style and processed via MathJax. Macros are to be avoided, of course.
\item A reference is introduced on one line in the form \newline
\texttt{<script language="javascript">bibref("Ramare*12")</script>}
\newline
 where Ramare*12 is the key of the bibtex entry, which has to be
 introduced in the Local-TME-EMT.bib file.  
\end{enumerate}

The file \texttt{Latex/booklet.tex} is the master file of the PDF
booklet and has to be edited by hand. The Perl script
\texttt{Biblio/UpDateBiblio.pl} will create tex-files that are to be
used as chapter for each html-file found in \texttt{Articles/}, save
the file \texttt{Template\_Article.html} of course.


\section{How to contribute}


Everyone is most welcome to help us keep track of the results. You can
do so by simply sending to the development team a mail with the proper
information (in bibtex format for the relevant part).  You may also
propose a new annoted bibliography, for instance for "Explicit results
in the combinatorial sieve", or any other missing entry.  There are
other ways to contribute, like modifying the CSS so that this site
would be readable under windows, a fact we do not guarantee. Or by
proposing a rewrite of some already present bibliography. Or anything
else we did not think about.


\section{Adding a part or a chapter}

Adding a chapter requires several steps.
\begin{itemize}
\item Modify the file \texttt{accueil.html}.
\item Modify in the file \texttt{../MiseEnPage.js} the variable
  \texttt{Architecture\_TME\_EMT}.
  This modification changes the numbers and each file
  \texttt{Articles/Art**.html} has to be subsequently modified at the
  level of the command \texttt{BandeauGeneral(2, "../../", [0, 2])}.
\item Modify accordingly the file \texttt{Latex/booklet.tex}.
\end{itemize}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter

\printbibliography[heading=bibintoc]




\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
